---
title: "Epigenomic Aging"
author: "Eric Kramer"
date: "1 June 2015"
output: rmarkdown::tufte_handout
---

```{r echo=F, include=FALSE}
library("dplyr")
library("tidyr")
library("knitr")
library("xtable")

options(xtable.comment = FALSE)
options(xtable.booktabs = TRUE)

load("../data/sample.info.Rdata")

```

# Introduction

DNA methylation is the binding of a methyl-group to guanine on DNA. It is well-known that DNA methylation affects gene expression, and that it is a strong contributor to differences in gene expression during tissue differentiation. More recent work has shown that DNA methylation is also affected by aging, socio-economic status and ancestral background.

# Assembly of DNA methylation database from GEO



I assembled a database of `r nrow(sample.info)` GEO samples of the Illumina 450k array. This database contains samples from `r length(unique(sample.info$tissue))` tissues, and ages ranging from neonates to `r max(sample.info$age, na.rm=T)` years old.




```{r, echo=FALSE, result='asis'}

tissue_counts = sample.info %>%
  group_by(tissue) %>%
  summarize(Count=n()) %>%
  select(Tissue=tissue, Count) %>%
  as.data.frame

kable(tissue_counts, caption="Number of samples for each tissue")

```

```{r, fig.width = 10, fig.height = 4, fig.cap = "Age distribution of GEO samples", echo=F, warning=FALSE, fig.fullwidth = TRUE}
library(ggplot2)
ggplot(sample.info,
       aes(age)) +
  geom_histogram(binwidth=2) +
  ylab("Count") +
  xlab("Age")
```


# Normalization 

BMIQ normalization was used on each sample. Downstream analysis found that this normalization minimized batch effects between different GEO entries.

```{r echo=FALSE, fig.cap="PCA of GEO samples, excluding outliers", fig.fullwidth = TRUE, fig.width = 10, fig.height = 4,}

load("./pcas/scores.no.lymphoblasts.Rdata")
ggplot(scores, aes(PC1, PC2, color=tissue)) +
  geom_point()

```


# Figures

## Margin Figures

Images and graphics play an integral role in Tufte's work. To place figures or tables in the margin you can use the `fig.margin` knitr chunk option. For example:

```{r, fig.margin = TRUE, fig.cap = "Sepal length vs. petal length, colored by species"}
library(ggplot2)
qplot(Sepal.Length, Petal.Length, data = iris, color = Species)
```

Note the use of the `fig.cap` chunk option to provide a figure caption. You can adjust the proportions of figures using the `fig.width` and `fig.height` chunk options. These are specified in inches, and will be automatically scaled down to fit within the handout margin.

## Full Width Figures

You can arrange for figures to span across the entire page by using the `fig.fullwidth` chunk option. 

```{r, fig.width = 10, fig.height = 2, fig.fullwidth = TRUE, fig.cap = "Full width figure"}
qplot(wt, mpg, data=mtcars, colour=factor(cyl))
```

Note the use of the `fig.width` and `fig.height` chunk options to establish the proportions of the figure. Full width figures look much better if their height is minimized.

## Main Column Figures

Besides margin and full width figures, you can of course also include figures constrained to the main column.

```{r, fig.cap = "Another figure"}
qplot(factor(cyl), mpg, data = mtcars, geom = "boxplot")
```

# Sidenotes

One of the most prominent and distinctive features of this style is the extensive use of sidenotes. There is a wide margin to provide ample room for sidenotes and small figures. Any use of a footnote will automatically be converted to a sidenote. ^[This is a sidenote that was entered using a footnote.] 

If you'd like to place ancillary information in the margin without the sidenote mark (the superscript number), you can use the `\marginnote` command. \marginnote{This is a margin note.  Notice that there isn't a number preceding the note.}

Note also that the two footnote references (`tufte_latex` and `books_be`, both defined below) were also included in the margin on the first page of this document.

# Tables

You can use the **xtable** package to format \LaTeX\ tables that integrate well with the rest of the Tufte handout style. Note that it's important to set the `xtable.comment` and `xtable.booktabs` options as shown below to ensure the table is formatted correctly for inclusion in the document.




[^tufte_latex]: https://code.google.com/p/tufte-latex/
[^books_be]: http://www.edwardtufte.com/tufte/books_be










